Publication Bias in Cochrane Meta-Analyses

### 2019-08-28 Presentation Final/present.Rnw
Deletion of unnecessary files

@@

adj.id.count <- length(unique(tp6$id))
adj.study.count <- length(unique(tp6$study.name))
adj.result.count <- dim(tp6)[1]
@

@@

\begin{frame}{Systematic Reviews}
\begin{itemize}
\item Summarise all evidence with regard to treatment with meta-analysis
\item Biased if non-significant results are not available and included
\end{itemize}
\end{frame}

@@

\vspace{-1mm}
$O = $ observed no. of significant results, $E$ expected based on power of studies.
\end{frame}

@@

\begin{frame}{Small Study Effect Tests}
\vspace{-4mm}
Weighted linear regression with std. error $x_i$ and effect size $y_i$:
\begin{align}
y_i &= \beta_0 + \beta_1 x_i + \epsilon, & \epsilon \sim N(0, x_i \sigma^2) \nonumber
\end{align}
Test for $H0: \beta_1 = 0$, no funnel plot asymmetry
\end{frame}

\begin{frame}{Adjustments for Binary Data}
As recommended by \citet{Sterne}
\begin{itemize}
\item Log odds ratio and risk ratio $\theta$ and standard error $\se_\theta$ are not independent
\item Use score of binomial likelihood at log odds ratio $\theta_{\textrm{H0}} = 0$ instead of 
log odds ratio and the inverse Fisher information instead of $\se_\theta$
\end{itemize}
\end{frame}

@@

p.dist.iv <- meta.iv %>%  ggplot(aes(x = pval1.egger)) + 
	geom_histogram(boundary = 0, binwidth = 0.1) + theme_bw() + 
  geom_text(data = dat_text, mapping = aes(x = 0.5, y = 24.5, label = label),  
  					color = "red", size = 4.5) + 
  ggtitle("Other Data (Egger)") + ylab("") +
	xlab(expression(paste(italic(p), "-value")))  + theme(
        plot.title = element_text(hjust = 0.5, face = "bold", size = 12, margin = margin(b = 20)),
        #axis.title.y = element_text(size=16, margin = margin(r = 10)), 
        axis.title.x = element_text(size=16, margin = margin(t = 10)), 
        axis.text= element_text(size=12),
        strip.text.x = element_text(size = 30))

@@

>>>>

bias.side.z <- function(yi, sei){
  alpha = 0.05
  to.ommit <- which(is.na(yi) | is.na(sei))

  if(length(to.ommit) > 0){
    yi <-  yi[-to.ommit]
    sei <- sei[-to.ommit]
  }

  to.ommit <- which(sei == 0)

  if(length(to.ommit) > 0){
    yi <-  yi[-to.ommit]
    sei <- sei[-to.ommit]
  }

  if(sum(pnorm(yi/sei) < .05) == sum(pnorm(yi/sei, lower.tail=FALSE) < .05)){
      side <- sign(est.fe <- rma(yi = yi, sei = sei, method = "FE")$b[1] )
  } else{
    side = ifelse(sum(pnorm(yi/sei) < .05) > sum(pnorm(yi/sei, lower.tail=FALSE) < .05),
                  -1, 1)
    }
  return(side)
}

@@

adj.id.count <- length(unique(tp6$id))
adj.study.count <- length(unique(tp6$study.name))
adj.result.count <- dim(tp6)[1]

#----------------------------------------------------------------------------------------------#
# missing.copas.histogram <- meta.f %>% ggplot(aes(x = missing.copas/k)) + geom_histogram(binwidth = 0.05, boundary = 0) + theme_bw() + xlab("Missing study fraction")
missing.copas.zeros <- length(which(meta.f$missing.copas[!is.na(meta.f$missing.copas)]/meta.f$k[!is.na(meta.f$missing.copas)] == 0))
missing.copas.range.frac <- c(missing.copas.zeros, quantile(x = meta.f$missing.copas[!is.na(meta.f$missing.copas)]/meta.f$k[!is.na(meta.f$missing.copas)], c(0.05, 0.25, 0.5, 0.75, 0.95)), mean(meta.f$missing.copas[!is.na(meta.f$missing.copas)]/meta.f$k[!is.na(meta.f$missing.copas)],na.rm = T) )
missing.copas.range.count <- c(missing.copas.zeros, quantile(x = meta.f$missing.copas[!is.na(meta.f$missing.copas)], c(0.05, 0.25, 0.5, 0.75, 0.95)), mean(na.rm = T,meta.f$missing.copas[!is.na(meta.f$missing.copas)])) 
missing.copas.table <- rbind(missing.copas.range.frac, missing.copas.range.count)
rownames(missing.copas.table) <- c("Overall fraction", "Missing number")
colnames(missing.copas.table) <- c("= 0", colnames(missing.copas.table)[2:6], "mean")
missing.copas.table <- missing.copas.table[c(2,1),]
@

@@

\begin{frame}{Systematic Reviews}
\begin{itemize}
\item Summarize all evidence with regard to treatment with meta-analysis
\item Biased if non-significant results are not available and included
\end{itemize}
\end{frame}

@@

\vspace{-1mm}
$O = $ observed no. of significant results, $E$ expected based on power of studies, $p$ = $E/n$.
\end{frame}

@@

\begin{frame}{Small Study Effect Tests}
\vspace{-4mm}
Weighted linear regression with std. error $x_i$ and effect size $y_i$:
\begin{align}
y_i &= \beta_0 + \beta_1 x_i + \epsilon_i, & \epsilon_i \sim N(0, x_i \sigma^2) \nonumber
\end{align}
Test for $H0: \beta_1 = 0$, no funnel plot asymmetry
\end{frame}

\begin{frame}{Adjustments for Binary Data}
As recommended by \citet{Sterne}
\begin{itemize}
\item Log odds ratio and risk ratio $\theta$ and standard error $\se_\theta$ are not independent
\item Use score of binomial likelihood at log odds ratio $\theta_{\textrm{H0}} = 0$ instead of 
log odds ratio, and the inverse Fisher information instead of $\se_\theta$
\end{itemize}
\end{frame}

@@

p.dist.iv <- meta.iv %>%  ggplot(aes(x = pval1.egger)) + 
	geom_histogram(boundary = 0, binwidth = 0.1) + theme_bw() + 
  geom_text(data = dat_text, mapping = aes(x = 0.5, y = 24.5, label = label),  
  					color = "red", size = 4.5) + 
  ggtitle("Other Data (Egger)") + ylab("") +
	xlab(expression(paste(italic(p), "-value")))  + theme(
        plot.title = element_text(hjust = 0.5, face = "bold", size = 12, margin = margin(b = 20)),
        axis.title.x = element_text(size=16, margin = margin(t = 10)), 
        axis.text= element_text(size=12),
        strip.text.x = element_text(size = 30))

@@

\begin{frame}{Adjustment}
Calculate unbiased estimates by
\begin{itemize}
\item Sensitivity analysis
\item Regression approach
\end{itemize}
\end{frame}

\begin{frame}{Selection Model}
\citet{Copas2}
\begin{align}
\theta_i &\sim N(\mu_i, \sigma_i^2) &
\mu_i \sim N(\theta, \tau^2) \nonumber
\end{align}
So we have $\theta_i = \mu_i + \sigma_i\epsilon_i \nonumber$. Introduce
\begin{itemize}
\item Baseline study retention rate $a$
\item Std. error dependent retention rate $b$
\item introduces correlation between $\epsilon_i$ and selection probability
\end{itemize}
%Model the selection process for increasing $a$ and $b$
\end{frame}

\begin{frame}{Sensitivity analysis}
To test a pair $a$,$b$,	include $\beta$:
\begin{align}
\theta_i &= \theta + \beta \se_i + \sigma_i \epsilon_i \nonumber
\end{align}
\vspace{-10mm}
\begin{itemize}
\item $\theta_i$ is the effect size of study $i$ and right-hand side is the fitted 
value of the model. 
\item Repeatedly test for $\beta_{\textrm{H0}} = 0$ %and estimate $\theta_{\textrm{Adj.}}$
\end{itemize}
\end{frame}

\begin{frame}{Regression}
Correct for small study effect \citep{Limitmeta}:
\begin{itemize}
\item Use global mean of regression $\beta_0$ from $y_i = \beta_0 + \beta_1 x_i + \epsilon_i, \epsilon_i \sim N(0, x_i \sigma^2) \nonumber$
\item Add minimal bias term $\beta_1 \cdot \tau$
\end{itemize}
$\theta_{\textrm{Adj.}} = \beta_0 + \beta_1\tau$
\end{frame}

\begin{frame}{Adjustment Methods}
Differences:
\begin{itemize}
\item Regression: uncertainty of bias parameter retained
\item Sensitivity analysis: Bias itself is not estimated, but most parsimonious scenario is chosen
\item Sensitivity analysis: No adjustment if no evidence for bias
\end{itemize}
Regression adjustment leads to higher uncertainty of $\theta_{\textrm{Adj.}}$
\end{frame}

\begin{frame}{Effect Measure Transformation}
\vspace{-3mm}
\begin{align}
d &= \theta \frac{\sqrt{3}}{\pi} & \se_d^2 =  \se^2\frac{\sqrt{3}}{\pi} \nonumber
\end{align}
\begin{align}
r &= \frac{d}{\sqrt{d^2 + a}} & a = (n_c + n_t)^2 / n_c n_t \nonumber
\end{align}
\begin{align}
z &= 0.5 \ln\bigg(\frac{1 + r}{1 - r}\bigg) & \nonumber
%z &= \arctan(r) & \nonumber
\se_z^2 &= \frac{1}{n-3}
\end{align}
\end{frame}

\begin{frame}[fragile]{Results}
<<echo = FALSE, cache = TRUE, fig.height = 4, message = FALSE, warning = FALSE>>=
diff.z.ranef.reg <- meta.f %>% 
  mutate(fixef = est.z.fixef, ranef = est.z.ranef,
         copas = est.z.copas, regression = est.z.reg) %>% 
  mutate(copas = case_when(sign(copas) == sign(ranef) ~  abs(copas),
                           sign(copas) != sign(ranef) ~ -abs(copas)),
         regression = case_when(sign(regression) == sign(ranef) ~  abs(regression),
                                sign(regression) != sign(ranef) ~ -abs(regression)),
         fixef = abs(fixef),
         ranef = abs(ranef)) %>% 
  select(fixef, ranef, copas, regression) %>% 
  gather(key = "method", value = "adjusted.z.score", regression) %>% 
  mutate(difference = ranef - adjusted.z.score) %>% 
  filter(difference > -.75 & difference < .75) %>% 
  ggplot(aes(x = difference)) + geom_histogram(binwidth = 0.1, center = 0) +
  theme_bw() + ggtitle("Regression") +
  xlab(expression(paste("Random effects ", italic(hat(z)), " - adjusted ", italic(hat(z))))) + 
	geom_vline(xintercept = 0, linetype = "dashed" ) + theme(
		axis.title.y = element_text(size=16, margin = margin(r = 10)), 
        plot.title = element_text(hjust = 0.5, face = "bold", size = 15, margin = margin(b = 22)),
        axis.title.x = element_text(size=14, margin = margin(t = 10)), 
        axis.text= element_text(size=12),
        strip.text.x = element_text(size = 30)) + xlim(c(-0.8, 0.7))
#--------------------------------------------------------------------------------------------------------------------#

diff.z.ranef.copas <- meta.f %>% 
  mutate(fixef = est.z.fixef, ranef = est.z.ranef,
         copas = est.z.copas, regression = est.z.reg) %>% 
  mutate(copas = case_when(sign(copas) == sign(ranef) ~  abs(copas),
                           sign(copas) != sign(ranef) ~ -abs(copas)),
         regression = case_when(sign(regression) == sign(ranef) ~  abs(regression),
                                sign(regression) != sign(ranef) ~ -abs(regression)),
         fixef = abs(fixef),
         ranef = abs(ranef)) %>% 
  select(fixef, ranef, copas, regression) %>% 
  gather(key = "method", value = "adjusted.z.score", copas) %>% 
  mutate(difference = ranef - adjusted.z.score) %>% 
  filter(difference > -.75 & difference < .75) %>% 
  ggplot(aes(x = difference)) + geom_histogram(binwidth = 0.1, center = 0) +
  theme_bw() + ylab("") + ggtitle("Selection Model") +
  xlab(expression(paste("Random effects ", italic(hat(z)), " - adjusted ", italic(hat(z))))) + 
  	geom_vline(xintercept = 0, linetype = "dashed" ) + theme(
        plot.title = element_text(hjust = 0.5, face = "bold", size = 15, margin = margin(b = 22)),
        axis.title.x = element_text(size=14, margin = margin(t = 10)), 
        axis.text= element_text(size=12),
        strip.text.x = element_text(size = 30)) + xlim(c(-0.8, 0.7))

grid.arrange(diff.z.ranef.reg, diff.z.ranef.copas, ncol = 2)
@
\end{frame}

\begin{frame}{Results}
% \vspace{-3mm}
% Compare wald test statistics (original effect size measures):
% 
% \vspace{-2mm}
<<echo = FALSE, warning = FALSE, cache = TRUE, fig.height = 4>>=
mid <- median(meta.f$pval.se)
method.names <- c(z.reg = "Regression", z.copas = "Selection Model")

decrease_text <- data.frame(label = c("decrease", "decrease"),
  method   = c("z.reg", "z.copas"))
increase_text <- data.frame(label = c("increase", "increase"),
                            method   = c("z.reg", "z.copas"))
meta.f %>% filter(z.reg > -40) %>% 
  select(pval.se, z.fixef, z.ranef, z.reg, z.copas) %>% 
  gather(key = "method", value = "test.stat", z.reg:z.copas) %>%
  ggplot(aes(x = test.stat, y = z.ranef, colour = pval.se)) + 
  facet_wrap(~ method, labeller = as_labeller(method.names)) + 
  scale_color_gradient2(midpoint=mid, low="firebrick4", mid = "gold",
                        high="chartreuse2", space ="Lab", name = expression(paste(italic(p),"-value"))) +
  geom_point(size = .6) + theme_bw() + guides(fill=guide_legend(title="p-value")) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, size = .3) +
  geom_abline(slope = 0, intercept = 0, linetype = 2, size = .3) +
  geom_text(data = decrease_text, mapping = aes(x = -11, y = 6, label = label), 
            color = "tan3") + 
  geom_text(data = decrease_text, mapping = aes(x = 11, y = -6, label = label), 
            color = "tan3") + 
  geom_text(data = increase_text, mapping = aes(x = 15, y = 4, label = label), 
            color = "chartreuse3") + 
  geom_text(data = increase_text, mapping = aes(x = -15, y = -4, label = label), 
            color = "chartreuse3") + 
  xlab(expression(paste("Adjusted ",  italic(z),  " test statistic"), sep = "")) + 
  ylab(expression(paste("Random effects z ", italic(z), " test statistic"), sep = "")) +
  theme(legend.title = element_text(size=14),
        axis.title.y = element_text(size=14, margin = margin(r = 20)), 
        axis.title.x = element_text(size=14, margin = margin(t = 20)), 
        legend.text = element_text(size =10), 
        axis.text= element_text(size=12),
        strip.text.x = element_text(size = 15),
        #plot.title = element_text(hjust = 0.5, face = "bold", size = 40, margin = margin(b = 20)),
  			strip.background = element_rect(fill = "white"))
@

\begin{frame}{Missing Study Number}
Missing study number according to selection model parameters: \Sexpr{format(round(sum(meta.f$missing.copas, na.rm = T)),big.mark = ",")}, \Sexpr{round(sum(meta.f$missing.copas, na.rm = T)/sum(meta.f$k),3)*100}\%

<<echo= FALSE, results='asis', cache = TRUE>>=
print(xtable(missing.copas.table, caption = "Fraction of missing studies and estimates of missing studies with their zero counts (``= 0''), quantiles and means.",
             label = "copas.missing", align = "lrrrrrrr", digits = c(0,0,0,0,1,1,1,1)),
      size = "footnotesize")
@
\end{frame}

\begin{frame}{Additional Results}
Linear mixed model with random effects for meta-analyses and reviews:

\begin{align}
\mathbf{y}_i \given U_j,U_k,\epsilon_{i} =  \beta_0 + x_i\beta_1 + U_j + U_k + \epsilon_{i} \nonumber
\end{align}

$\rightarrow$ Increased power to estimate $\beta$
\end{frame}

\begin{frame}{Mixed Linear Model}
AIC improvement to null fit: 1820 to 1306. \\
Bias parameter $\beta_1$ = 0.52 (95\%CI: 0.44,0.59)

\vspace{-12mm}
\begin{figure}
<<echo = FALSE, warning=FALSE, cache = TRUE, fig.height = 4>>=
reg.dat <- data.ext2 %>% filter(meta.id %in% meta.f$meta.id) #Filter meta-analysis data
reg.dat <- reg.dat %>% group_by(meta.id) %>%
  filter(var.z != 0) %>% filter(!is.na(z)) %>% filter(!is.na(var.z)) %>%
  mutate(se.z.mirrored = sqrt(var.z)) %>% filter(!is.na(se.z.mirrored)) %>%
  mutate(bias.side.z = bias.side.z(yi = z, sei = se.z.mirrored),
  z.mirrored = z * bias.side.z) #Mirror effect sizes on one side
reg.dat <- reg.dat %>%  filter(!is.na(z.mirrored)) %>%
  filter(!is.na(se.z.mirrored))
reg.dat <- reg.dat %>% filter(se.z.mirrored < 3)
publication.bias.fit <- lme(fixed = z.mirrored ~ se.z.mirrored, 
														random = list(~ 1 | id, ~ 1 | meta.id), 
														weights = ~ se.z.mirrored^2, method = "ML", data = reg.dat)
publication.bias.rs.fit <- lme(fixed = z.mirrored ~ se.z.mirrored,
                          random = list(~ 1 | id, ~ meta.id | meta.id),
                       weights = ~ se.z.mirrored^2, method = "ML", data = reg.dat)
plot(y = reg.dat$z.mirrored, x = reg.dat$se.z.mirrored, cex = .1, cex.lab = 1.3,
		 xlab = "standard error", ylab = "z mirrored")
publication.bias.fit.coef <- summary(publication.bias.fit)$coefficients$fixed
abline(a = publication.bias.fit.coef, col = "red", lty = 2, lwd = 1.5)
publication.bias.rs.fit.coef <- summary(publication.bias.rs.fit)$coefficients$fixed
abline(a = publication.bias.rs.fit.coef, col = "blue", lty = 3, lwd = 1.5)
@
\end{figure}
\end{frame}

\begin{frame}{Limitations}
Small study effect $\neq$ publication bias. Also:
\begin{itemize}
\item True heterogeneity
\item Selective outcome reporting
\item Delayed publication
\item ...
\end{itemize}
\vspace{-3mm}
Small study effect tests do not take into account significance directly.
\end{frame}

\begin{frame}{Limitations}
\begin{itemize}
\item Unknown amount of secondary outcomes
\item Small amount of adverse outcomes
\item Publication bias unknown for excluded data
\item Exploratory Analysis!
\end{itemize}
\end{frame}

\begin{frame}{Implications}
\begin{itemize}
\item Results are largely in line with previous research
\item Underline the need for examination of publication and other biases in meta-analyses
\item Otherwise, validity of meta-analyses can be contested
\end{itemize}
\end{frame}

\begin{frame}
Backup slides:
\end{frame}

\begin{frame}[fragile]{Transformation}

\vspace{-6mm}
<<echo = FALSE, warning=FALSE, fig.height=4.8, cache = TRUE>>=
par(mfrow = c(2,3), cex.lab = 1.5, cex.axis = 1.5)
example <- data.ext2 %>% filter(meta.id == 81244)
funnel(limitmeta(metagen(TE = effect, se = se, studlab = study.name, sm = "MD", data = example)), col.adjust = "blue", col.line = "blue")
funnel(limitmeta(metagen(TE = smd.pool, se = se.smd.pool, studlab = study.name, sm = "SMD", data = example)), col.adjust = "blue", col.line = "blue")
funnel(limitmeta(metagen(TE = z, seTE = sqrt(var.z), studlab = study.name, sm = "ZCOR", data = example)), col.adjust = "blue", col.line = "blue")

example <- data.ext2 %>% filter(meta.id == 32532)
funnel(limitmeta(metagen(TE = log(effect), se = se, studlab = study.name, sm = "RR", data = example)), col.adjust = "blue", col.line = "blue")
funnel(limitmeta(metagen(TE = smd.pool, se = se.smd.pool, studlab = study.name, sm = "SMD", data = example)), col.adjust = "blue", col.line = "blue")
funnel(limitmeta(metagen(TE = z, seTE = sqrt(var.z), studlab = study.name, sm = "ZCOR", data = example)), col.adjust = "blue", col.line = "blue")
@
\end{frame}
